---
title: "Final"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tsdl)
tsdl_sales<-subset(tsdl,"Sales")
attributes(tsdl_sales[[1]])
tsdl_sales[[1]]
original<-tsdl_sales[[1]]
sale <- tsdl_sales[[1]][c(-187:-192)]
#sale<-tsdl_sales[[1]]
sale.ts <- ts(sale, start = c(1960,1), frequency = 12)
sale.ts
ts.plot(sale.ts,ylab = "Gasoline (Million Gallons)",  main = "Monthly Gasoline demand vs. Time")
```
```{r}
# Plot has trend and seasonality components; we must remove them
library(MASS)
t = 1:length(sale.ts)
fit = lm(sale.ts ~ t)
bcTransform = boxcox(sale.ts ~ t, plotit = TRUE)
```
```{r}
#Box Cox transform does not include lamba = 0 in CI, so \dfrac{y_i^{\lambda}-1}{\lambda}

lambda = bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
sale.bc = (1/lambda)*(sale^lambda-1)
sale.log = log(sale.ts)
sale.sqrt = sqrt(sale.ts)
op <- par(mfrow = c(1,2))
ts.plot(sale.ts,main = "Original data",ylab = expression(X[t]))
ts.plot(sale.bc,main = "Box-Cox tranformed data", ylab = expression(Y[t]))
ts.plot(sale.ts,main = "Original data",ylab = expression(X[t]))
ts.plot(sale.log,main = "Log tranformed data", ylab = expression(Y[t]))
var(sale.ts)
var(sale.bc)
var(sale.log)
var(sale.sqrt)
```
Log transformation has the lowest variance, so we go with log transformation.
```{r}
# Plot has trend and seasonality components; we must remove them
op = par(mfrow = c(1,2))
acf(sale.log,lag.max = 60,main = "")
pacf(sale.log,lag.max = 60,main = "")
title("Log Transformed Time Series", line = -1, outer=TRUE)

x12 = diff(sale.ts, 12)
var(x12)
ts.plot(x12)
op = par(mfrow = c(1,2))
acf(x12,lag.max = 60,main = "")
pacf(x12,lag.max = 60,main = "")
title("Differenced at 12", line = -1, outer=TRUE)
x12diff1 = diff(x12, 1)
var(x12diff1)
ts.plot(x12diff1)
op = par(mfrow = c(1,2))
acf(x12diff1,lag.max = 60,main = "")
pacf(x12diff1,lag.max = 60,main = "")
title("x12diff1", line = -1, outer=TRUE)
x1 = diff(sale.ts, 1)
var(x1)
ts.plot(x1)
op = par(mfrow = c(1,2))
acf(x1,lag.max = 60,main = "")
pacf(x1,lag.max = 60,main = "")
title("Differenced at 1", line = -1, outer=TRUE)
x1diff12 = diff(x1, 12)
var(x1diff12)
ts.plot(x1diff12)
op = par(mfrow = c(1,2))
acf(x1diff12,lag.max = 60,main = "")
pacf(x1diff12,lag.max = 60,main = "")
title("x1diff12", line = -1, outer=TRUE)


y12 = diff(sale.log, 12)
var(y12)
ts.plot(y12)
op = par(mfrow = c(1,2))
acf(y12,lag.max = 60,main = "")
pacf(y12,lag.max = 60,main = "")
title("Differenced at 12 Log Transformed Time Series", line = -1, outer=TRUE)

y1 = diff(sale.log, 1)
var(y1)
ts.plot(y1)
op = par(mfrow = c(1,2))
acf(y1,lag.max = 60,main = "")
pacf(y1,lag.max = 60,main = "")
title("Differenced at 1 Log Transformed Time Series", line = -1, outer=TRUE)

y12diff12 = diff(y12, 12)
var(y12diff12)
ts.plot(y12diff12)

y12diff1 = diff(y12, 1)
var(y12diff1)
ts.plot(y12diff1)
op = par(mfrow = c(1,2))
acf(y12diff1,lag.max = 60,main = "")
pacf(y12diff1,lag.max = 60,main = "")
title("y12diff1", line = -1, outer=TRUE)

y1diff12 = diff(y1, 12)
var(y1diff12)
a1 = diff(y1diff12, 1)
var(a1)
a12 = diff(y1diff12, 12)
var(a12)
ts.plot(y1diff12)
op = par(mfrow = c(1,2))
acf(y1diff12,lag.max = 60,main = "")
pacf(y1diff12,lag.max = 60,main = "")
title("y1diff12", line = -1, outer=TRUE)

y1diff1 = diff(y1, 1)
var(y1diff1)
ts.plot(y1diff1)
```
```{r}
op = par(mfrow = c(1,2))
acf(y1diff12,lag.max = 60,main = "")
pacf(y1diff12,lag.max = 60,main = "")
title("De-trended/seasonalized Time Series", line = -1, outer=TRUE)
```
ACF cuts off after lag 1 and possibly lag 2 for every 12 ticks while the PACF tails off leading me to believe that the model follows a (0,1,1) or a (0,1,2) seasonal component. For the non seasonal part, PACF cuts off after lag = 2 and lag = 5 meaning that it probably has an AR(2) or AR(5) counterpart.


```{r}
library(astsa)
m1<-sarima(sale.ts,2,1,0,0,1,1, 12)

m2<-sarima(sale.ts,2,1,1,0,1,1, 12)

m3<-sarima(sale.ts,2,1,2,0,1,1, 12)

m4<-sarima(sale.ts,2,1,3,0,1,1, 12)

m5<-sarima(sale.ts,5,1,1,0,1,1, 12)

m6<-sarima(sale.ts,5,1,0,0,1,1, 12)

m7<-sarima(sale.ts,5,1,2,0,1,1, 12)

m8<-sarima(sale.ts,5,1,3,0,1,1, 12)

m9<-sarima(sale.ts,2,1,0,0,1,2, 12)

m10<-sarima(sale.ts,2,1,1,0,1,2, 12)

m11<-sarima(sale.ts,2,1,2,0,1,2, 12)

m12<-sarima(sale.ts,2,1,3,0,1,2, 12)

m13<-sarima(sale.ts,5,1,1,0,1,2, 12)

m14<-sarima(sale.ts,5,1,0,0,1,2, 12)

m15<-sarima(sale.ts,5,1,2,0,1,2, 12)

m16<-sarima(sale.ts,5,1,3,0,1,2, 12)

m1$AIC
m2$AIC
m3$AIC
m4$AIC
m5$AIC
m6$AIC
m7$AIC
m8$AIC
m9$AIC
m10$AIC
m11$AIC
m12$AIC
m13$AIC
m14$AIC
m15$AIC
m16$AIC
```
The lowest AIC scores were 18.95 for model 16, 18.97 for model 15, 18.896 for model 12, 18.956 for model 8, 18.983 for model 7 and 18.984 for model 4. By the law of parsimony, we want to choose the model with the fewest parameters, so we decide on model 4 with 6 parameters.
```{r}
fit<-ar(y12diff1, method=c("yule-walker"))
fit
fit2<-ar(sale.ts, method=c("yule-walker"))
fit2
var(y12)
library(qpcR)
aiccs<-matrix(NA, nr = 6, nc = 6)
dimnames(aiccs) = list(p = 0:5, q = 0:5)
for(p in 0:5)
{
  for(q in 0:5)
  {
    aiccs[p+1, q+1] = AICc(arima(y1diff12, order=c(p,1,q),seasonal=list(order=c(0,1,1),period=12), method = "ML"))
  }
}
aiccs
```
```{r}
model4<- arima(sale.ts, order = c(2, 1, 3), seasonal = list(order = c(0, 1, 1), period = 12))
model4
Box.test(model4$residuals, lag = 14, type = c("Box-Pierce"), fitdf = 6)
Box.test(model4$residuals, lag =14, type = c("Ljung-Box"), fitdf = 6)
Box.test((model4$residuals)^2, lag =14, type = c("Ljung-Box"))
shapiro.test(model4$residuals)

ts.plot(model4$residuals, main="Model Residuals")
op = par(mfrow = c(1,2))
acf(model4$residuals,main = "Autocorrelation")
pacf(model4$residuals,main = "Partial Autocorrelation")
hist(model4$residuals,main = "Histogram")
qqnorm(model4$residuals)
qqline(model4$residuals, col = "blue")

model16<- arima(sale.ts, order = c(5, 1, 3), seasonal = list(order = c(0, 1, 2), period = 12))
model16
Box.test(model16$residuals, lag = 14, type = c("Box-Pierce"), fitdf = 10)
Box.test(model16$residuals, lag =14, type = c("Ljung-Box"), fitdf = 10)
Box.test((model16$residuals)^2, lag =14, type = c("Ljung-Box"), fitdf = 0)
shapiro.test(model16$residuals)

ts.plot(model16$residuals, main="Model Residuals")
op = par(mfrow = c(1,2))
acf(model16$residuals,main = "Autocorrelation")
pacf(model16$residuals,main = "Partial Autocorrelation")
hist(model16$residuals,main = "Histogram")
qqnorm(model16$residuals)
qqline(model16$residuals, col = "blue")
```
```{r}
plot.roots <- function(ar.roots=NULL, ma.roots=NULL, size=2, angles=FALSE, special=NULL, sqecial=NULL,my.pch=1,first.col="blue",second.col="red",main=NULL)
{xylims <- c(-size,size)
      omegas <- seq(0,2*pi,pi/500)
      temp <- exp(complex(real=rep(0,length(omegas)),imag=omegas))
      plot(Re(temp),Im(temp),typ="l",xlab="x",ylab="y",xlim=xylims,ylim=xylims,main=main)
      abline(v=0,lty="dotted")
      abline(h=0,lty="dotted")
      if(!is.null(ar.roots))
        {
          points(Re(1/ar.roots),Im(1/ar.roots),col=first.col,pch=my.pch)
          points(Re(ar.roots),Im(ar.roots),col=second.col,pch=my.pch)
        }
      if(!is.null(ma.roots))
        {
          points(Re(1/ma.roots),Im(1/ma.roots),pch="*",cex=1.5,col=first.col)
          points(Re(ma.roots),Im(ma.roots),pch="*",cex=1.5,col=second.col)
        }
      if(angles)
        {
          if(!is.null(ar.roots))
            {
              abline(a=0,b=Im(ar.roots[1])/Re(ar.roots[1]),lty="dotted")
              abline(a=0,b=Im(ar.roots[2])/Re(ar.roots[2]),lty="dotted")
            }
          if(!is.null(ma.roots))
            {
              sapply(1:length(ma.roots), function(j) abline(a=0,b=Im(ma.roots[j])/Re(ma.roots[j]),lty="dotted"))
            }
        }
      if(!is.null(special))
        {
          lines(Re(special),Im(special),lwd=2)
        }
      if(!is.null(sqecial))
        {
          lines(Re(sqecial),Im(sqecial),lwd=2)
        }
        }

model<- arima(sale.ts, order = c(2, 1, 3), seasonal = list(order = c(0, 1, 1), period = 12))
model
plot.roots(NULL,polyroot(c(1, 1.0298, 0.8367)), main="roots of ar part")
plot.roots(NULL,polyroot(c(1, -0.0118, 0.0217, -0.3765)), main="roots of ma part")
plot.roots(NULL,polyroot(c(1, -0.6736)), main="roots of seasonal ma part")
```
All the roots are outside the unit circle, this passes the root test, meaning the model is causal and invertible.
All the coefficients are significant. 
```{r}
sarima.for(sale.ts, 6, 2, 1, 3, 0, 1, 1, 12)
x<-c(1,2,3,4,5,6)
Observed <- c(255918, 244642, 237579, 237579, 217775, 227621)
Upper<-c(239821.816, 248806.544, 214537.774, 225749.588, 205510.49, 211875.78)
Lower<-c(261786.184, 270789.856, 237027.226, 250278.212, 230039.51, 237096.02)
Predicted<-c(250804.0, 259798.2, 225782.5, 238013.9, 226507.4, 224485.9)
plot(x,Observed, xlab = "Time", ylab="Gasoline Demand", main="Predicted vs Observed", pch=15, col="black")
lines(x, Upper, pch=16, col="blue")
lines(x, Lower, pch=16, col="blue")
points(x, Predicted, pch=16, col="red")
```